Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-13 talk/icooolps2011/code/map.tex
replace "pure" with "invariant", add a precise definition of the term

    @purefunction
    def getindex(self, name):
        return self.indexes.get(name, -1)

    @purefunction
    def add_attribute(self, name):
        if name not in self.other_maps:
            newmap = Map()
            newmap.indexes.update(self.indexes)
            newmap.indexes[name] = len(self.indexes)
            self.other_maps[name] = newmap
        return self.other_maps[name]

>>>>

    @invariant
    def getindex(self, name):
        return self.indexes.get(name, -1)

    @invariant
    def add_attribute(self, name):
        if name not in self.other_maps:
            newmap = Map()
            newmap.indexes.update(self.indexes)
            newmap.indexes[name] = len(self.indexes)
            self.other_maps[name] = newmap
        return self.other_maps[name]

### 2011-04-13 talk/icooolps2011/paper.tex
replace "pure" with "invariant", add a precise definition of the term

In this paper we present two of these hints that are extensively used in the
PyPy project to improve the performance of its Python interpreter.

@@

\cfbolz{XXX kill the next paragraph? the info is repeated in the list below}
Concretely these hints are used to control how the optimizer of the
tracing JIT can improve the traces of the object model. In particular the hints
influence the constant folding
optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constant by feeding back runtime values. The
second hint allows the definition of additional foldable operations.

@@

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace by
 feeding back runtime information into compilation.
 \item A way to annotate operations as pure which the constant folding
 optimization then recognizes.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
\end{itemize}

@@

\cfbolz{XXX stress more that "the crux of the techniques and a significant
portion of new contributions in the paper are from how to refactoring codes to
expose likely runtime constants and pure functions"}

@@

\begin{itemize}
    \item the arguments of an operation actually need to all be constant,
    i.e. statically known by the optimizer
    \item the operation needs to be \emph{pure}, i.e. always yield the same result given
    the same arguments.
\end{itemize}

@@

\subsection{Declaring New Pure Operations}

In the previous section we saw a way to turn arbitrary variables into constants. All
pure operations on these constants can be constant-folded. This works well for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not pure and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    @purefunction
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

\cfbolz{XXX define the meaning of purefunction more precisely, particularly because add\_attribute has side effects, which is confusing}

\cfbolz{should we mention that pure functions are not actually called by the optimizer, but the values that are seen during tracing are used?}

@@

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to mess things up. If a
function is annotated to be pure, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{purefunction}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}.

\cfbolz{XXX mention a possible debug mode for findings bugs in this, or too many values in promotion. stress more that promote is safe}

\subsubsection{Observably Pure Functions}

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is a pure function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{purefunction} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If this function is analyzed, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function from the outside is pure. This is
a property that is not easily detectable by analysis. Therefore, the purity
of this function needs to be annotated manually.

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two previous sections don't seem to help with the current object model. There is
no pure function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

This is a common problem when trying to apply hints. Often, the interpreter
needs a small rewrite to expose the pure functions and nearly-constant objects
that are implicitly there. In the case of instance fields this rewrite is not
entirely obvious. The basic idea is as follows. In theory instances can have
arbitrary fields. In practice however many instances share their layout (i.e.
their set of keys) with many other instances.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Therefore they have to be immutable, which means
that their \texttt{getindex} method is a pure function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also pure,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

@@

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a pure function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

What we would really like is if the \texttt{Class.find\_method} method were pure.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
pure, because when the version object changes, the result of the call might be
different than the previous one.

@@

XXX Peng Wu and David Edelsohn

>>>>

In this paper we present two of these hints that are extensively used in the
PyPy project to improve the performance of its Python interpreter, particularly
of the object model.

@@

\cfbolz{XXX kill the next paragraph? the info is repeated in the list below}
In particular the hints
influence the constant folding
optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constant by feeding back runtime values. The
second hint allows the definition of additional foldable operations.

@@

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace by
 feeding back runtime information into compilation.
 \item A way to annotate operations which the constant folding
 optimization then recognizes and exploits.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
\end{itemize}

@@

\cfbolz{XXX stress more that "the crux of the techniques and a significant
portion of new contributions in the paper are from how to refactoring codes to
expose likely runtime constants and invariant functions"}

@@

\begin{itemize}
    \item the arguments of an operation actually need to all be constant,
    i.e. statically known by the optimizer
    \item the operation needs to be \emph{constant-foldable}, i.e. always yield
    the same result given the same arguments.
\end{itemize}

@@

\subsection{Declaring New Foldable Operations}

In the previous section we saw a way to turn arbitrary variables into constants. All
foldable operations on these constants can be constant-folded. This works well for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not foldable and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

This hint can be used to mark functions as \emph{invariant}. A function is
termed invariant if, during the execution of the program, successive calls to
the function with identical arguments always return the same result. From this
definition follows that a call to an invariant function with constant arguments
in a trace can be replaced with the result of the call.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is an invariant function. To communicate this, there is a \texttt{invariant} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

    @invariant
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}

@@

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as invariant, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

@@

On the one hand, the \texttt{invariant} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
function is annotated to be invariant, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{invariant}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}. We hope to introduce a
debugging mode which would (slowly) check whether the annotation is applied
incorrectly to mitigate this problem.

\subsubsection{Observably invariant Functions}

\cfbolz{XXX do we kill this section?}

Why can't we simply write an analysis to find out that the \texttt{x} fields of the
\texttt{A} instances is immutable and deduce that \texttt{compute} is an invariant function,
since it only reads the \texttt{x} field and does not have side effects? This might
be possible in this particular case, but in practice the functions that are
annotated with the \texttt{invariant} decorator are usually more complex.
The easiest example for this is that of a function that uses memoization to
cache its results. If this function is analyzed, it looks like the function has
side effects, because it changes the memoizing dictionary. However, because this side
effect is not externally visible, the function is still invariant. This is
a property that is not easily detectable by analysis.

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints we have looked at in the
two previous sections don't seem to help with the current object model. There is
no invariant function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

This is a common problem when trying to apply hints. Often, the interpreter
needs a small rewrite to expose the invariant functions and nearly-constant objects
that are implicitly there. In the case of instance fields this rewrite is not
entirely obvious. The basic idea is as follows. In theory instances can have
arbitrary fields. In practice however many instances share their layout (i.e.
their set of keys) with many other instances.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Therefore they have to be immutable, which means
that their \texttt{getindex} method is an invariant function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also invariant,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

@@

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
an invariant function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

What we would really like is if the \texttt{Class.find\_method} method were invariant.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is an invariant operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@invariant}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
invariant, because when the version object changes, the result of the call might be
different than the previous one.

@@

XXX Peng Wu and David Edelsohn, Laura Creighton
